---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE)
library(tidyverse)
library(scales)
library(lubridate)
source("scripts/viridis.R")
```

## Data

The analysis involve five data sets involving international flights arriving to Chicago's O'Hare airport from January 1, 2016 through June 30, 2020 with one separate file for each year.

Each data set is in five separate CSV files:
`ORD-2016.csv`, `ORD-2017.csv`, `ORD-2018.csv`, `ORD-2019.csv`, and `ORD-2020.csv`.

> Read in the five data sets. 
Add columns for *year*, *month* (character valued, Jan-Dec),
*day* (day of the month), and *wday* (day of the week, character valued, Sun - Sat).
Reorder the variables so that these new variables all appear directly after date.
Remove the terminal variable.
Rename *all_total* to *passengers*, *all_flights* to *flights*, and *all_booths* to *booths*.
Arrange the rows by date and hour.

```{r Read Data}
ord_2016 <- read_csv("data/ORD-2016.csv") %>%
  mutate(date = mdy(date)) 

ord_2017 <- read_csv("data/ORD-2017.csv") %>%
  mutate(date = mdy(date)) 

ord_2018 <- read_csv("data/ORD-2018.csv") %>%
  mutate(date = mdy(date)) 

ord_2019 <- read_csv("data/ORD-2019.csv") %>%
  mutate(date = ymd(date)) 

ord_2020 <- read_csv("data/ORD-2020.csv") %>%
  mutate(date = ymd(date)) 

data <- bind_rows(ord_2016, ord_2017, 
                  ord_2018, ord_2019, ord_2020) %>%
  select(-terminal) %>%
  mutate(year = year(date)) %>%
  mutate(month = month(date, label = T)) %>%
  mutate(day = day(date)) %>%
  mutate(wday = wday(date, label = T)) %>%
  rename(passengers = all_total,
        flights = all_flights,
         booths = all_booths) %>%
  select(airport, date, year, month, day, wday, everything()) %>%
  arrange(date, hour)

rm(ord_2016)
rm(ord_2017)
rm(ord_2018)
rm(ord_2019)
rm(ord_2020)
```

## Analysis

> Check if there is any missing data
Check if there are missing dates from January 1, 2016 through June 30, 2020

```{r Check Missing}
missing_rows <- data %>%
  drop_na() %>%
  summarise(count = 29450 - n())

all_dates = tibble(
  date = seq(ymd("2016-01-01"), ymd("2020-06-30"), 1))

missing <- full_join(all_dates, select(data, date, flights)) %>%
  filter(is.na(flights)) %>%
  select(date)

rm(all_dates)
missing_rows
missing
```
### Conclusions:

No rows contain missing data.

There are 8 dates missing, which are:

- 03/07/2016
- 03/08/2016
- 03/09/2016
- 05/08/2018
- 10/28/2019
- 10/29/2019
- 10/30/2019
- 02/08/2020



> Calculate the total numbers of flights and passengers in each month and year and store this information in a table.
Summarize this table to find the total number of passengers and flights in each year from 2016 - 2019.

```{r Month-Year Summarization, message = FALSE}
total_year <- data %>%
  group_by(year, month) %>%
  summarise(total_flights = sum(flights),
            total_passengers = sum(passengers))

highest <- total_year %>%
  group_by(year) %>%
  summarise(flights = sum(total_flights),
            passengers = sum(total_passengers))

total_year
```
###Conclusions:

Year with highest total flights: 2018

Year with highest total passengers: 2019



>  Display the total number of passengers by month and year with a bar chart where month is the primary variable on the x-axis and there is a separate bar (not stacked, and filled with a different color) for each year.

```{r Display}
ggplot(total_year) +
  geom_col(mapping = aes(x=month, y=total_passengers, fill=as.character(year)), position = position_dodge2(preserve = "single")) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_discrete() +
  labs(fill = "Year") +
  xlab("Months") +
  ylab("Total passengers") +
  ggtitle("Total Passengers per Months")
```

### Conclusions:

The overall trend of past 4 years excluding 2020 is that the total number of passengers increases the beginning of the year and peak at July, and then decrease until the end of the year. The total number of passengers in each months for each year also increase as the year increases. Besides, that of 2020 does not follow the trend above as the number of passengers from the beginning of the year decreases sharply until now, and the total number of passengers of 2020 for months until now are all less then the that of the previous years (except for January when the number is close to 2019). For the past 3 months, the number of passengers are less than 1/10 of the previous years.



> Add a `weekend` column to the combined data set which is TRUE for Saturdays and Sundays and FALSE for other days.
Make a scatter plot with the average time for US passengers on the x-axis and the average time for non-US passengers on the y-axis.
Use different colors for weekend and weekdays.
Add straight regression lines to the plot, separately for weekends and weekdays
Plot the data from different years in different facets.

```{r Weekend Summarization}
data_weekend <- data %>%
  mutate(weekend = wday %in% c("Sat", "Sun"))

ggplot(data_weekend, aes(x=us_avg_wait, y = non_us_avg_wait, color = weekend)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  geom_smooth(method = "lm") +
  facet_wrap(~year) +
  scale_color_discrete(labels = c("Weekday", "Weekend"), name = "Day of Week") +
  xlab("Average time for US passengers") +
  ylab("Average time for non-US passengers")

```

### Conclusions:

The pattern is that the average wait time of non us and us flyers are positively correlated

The average wait time for non us flyers are always higher than us flyers (since the slope is greater than 1 for both weekday and weekend). 
Back in 2016, the ratio between the average wait time of non us and us flyers are higher in weekday than in weekend. However, the gap is narrowed by time and in 2020, they are almost the same.



>Calculate separately for each year, the fraction of cases (a case is a single hour on a single date) for which the average time to get through passport control is greater for non US passport holders than it is for passport holders. Comment on how these values relate to the graphs in the previous problem.

```{r US and Non-US Comparison}
q6 <- data %>%
  group_by(year) %>%
  summarise(sum = sum(us_avg_wait < non_us_avg_wait)/n())

q6
```

### Conclusion:

Around 90% of the cases the non us flyers have to wait longer than the us flyers, which matches with the conclusions drawn from the graph above.



> Add a column named `booth_rate` to the data set which estimates the average number of passengers per booth per hour. For example, if 1000 passengers arrive between 05:00 and 06:00, the average wait time is 40 minutes, and there are 10 booths open, then an estimate of the total number of passengers per booth per hour could be computed like this: 1000/10 = 100 passengers per booth; (40 minutes per passenger * 1 hour per 60 minutes) = 2/3 hours per passenger; booth rate = 100 / (2/3) = (1000 * 60) / (10 * 40) = 150 passengers per booth per hour. This is an estimate because it assumes available booths change on the hour and it ignores how rates change when passenger wait durations stretch into the next time period.
Add another column called `time_of_day` which takes the value "overnight" from 1am to 5am, "early morning" from 5am to 8am, "morning" from 8am to noon, "afternoon" from noon to 5pm, and "early evening" from 5pm to 8pm, and "late evening" from 8pm to 1am.
Use `reorder()` to put the `time_of_day` variable in this order.

> Filter out cases where there are fewer than 200 total passengers, the average wait time is zero, or the booth rate is over 500. Make side_by_side boxplots of the booth rate versus the day of the week using different colors for each day of the week, different facets for each time of day, and fill color white if it is a weekday and gray if it is on the weekend.

```{r Booth Rate}
q7 <- data_weekend %>%
  mutate(booth_rate = passengers/booths/all_avg_wait*60) %>%
  separate(hour, into = c("start", "end"), remove=F) %>%
  separate(start, into = c("start", "minute"), sep = 2) %>%
  mutate(start = as.numeric(start)) %>%
  select(-minute) %>%
  separate(end, into = c("end", "minute"), sep = 2) %>%
  select(-minute) %>%
  mutate(end = as.numeric(end)) %>%
  mutate(time_of_day = case_when(
    (start >= 1 & end <= 5 & end != 0) ~ "overnight",
    start >= 5 & end <=8 & end != 0 ~ "early morning",
    start >= 8 & end <= 12 & end != 0 ~ "morning",
    start >= 12 & end <= 17 & end != 0 ~ "afternoon",
    start >= 17 &end <= 20 & end != 0 ~ "early evening",
    TRUE ~ "late evening"
  )) %>%
  mutate(order = case_when(
    time_of_day == "overnight" ~ 1,
    time_of_day == "early morning" ~2,
    time_of_day == "morning" ~ 3,
    time_of_day == "afternoon" ~ 4,
    time_of_day == "early evening" ~ 5,
    time_of_day == "late evening" ~ 6,
  )) %>%
  mutate(time_of_day = reorder(time_of_day, order)) %>%
  filter(!(passengers < 200 | all_avg_wait == 0 | booth_rate > 500))

ggplot(q7) +
  geom_boxplot(mapping = aes(y = booth_rate, x = wday, color = wday, fill = weekend)) +
  scale_fill_manual(values=c("white", "gray"), labels = c("Weekday", "Weekend")) +
  facet_wrap(~time_of_day) +
  scale_color_discrete(name = "Day of Week") +
  ylab("Booth rate") +
  theme(axis.text.x = element_text(angle = 90, hjust=1))
```

### Conclusions:

- Early morning has the lowest booth rate
- Booth rates tend to be higher on the weekend during the afternoon and early evening, otherwise, the booth rates of weekend and weekdays are close to each other for the other time of the day
- This effect is quite small in comparison to the variation in the booth rate within a day of week and time of day